Multi-resolution Neural Field Estimation

### 2011-03-07 Multiresolution_framework/ltx/manuscript.tex
changed figure name to IEEE format
edits to the whole paper

Equations \eqref{eq:KernelExpansion} and \eqref{eq:FieldExpansion} are infinite series expansions and must be truncated at some level $j$. The field and the connectivity kernel are approximated by
\begin{align}
	w\left(r-r'\right) &\approx \boldsymbol\theta^\top\boldsymbol\lambda\left(r-r'\right) 
	\label{eq:KernelFiniteExpansion} \\
	v_t\left(r\right) &\approx \boldsymbol\mu\left(r\right)^\top\mathbf{x}_t,
	\label{eq:FieldFiniteExpansion}
\end{align}
where the unknown parameter and state vectors, $\boldsymbol\theta \in \mathbb{R}^{n_{\theta}}$ and $\mathbf{x}_t \in \mathbb{R}^{n_x}$, are defined as 
\begin{align}
\boldsymbol\theta^\top &=\left[\begin{array}{ccccc} \boldsymbol\alpha_{j_0}^\top & \boldsymbol\beta_{j_0}^\top & \boldsymbol\beta_{j_0+1}^\top & \cdots & \boldsymbol\beta_{j}^\top \end{array}\right] 
\label{KernelWeights} \\
\mathbf{x}_{t}^\top &=\left[\begin{array}{ccccc}\mathbf{x}_{t,j_{0}}^\top &  \check{\mathbf{x}}_{t,j_{0}}^\top & \check{\mathbf{x}}_{t,j_{0}+1}^\top & \cdots & \check{\mathbf{x}}_{t,j}^\top\end{array}\right].
\label{FieldWeights}
\end{align}
The kernel and field approximation coefficient vectors, $\boldsymbol \alpha_{j_0}$ and $\mathbf{x}_{t,j_{0}}$, contain all the coefficients $\left\lbrace\alpha_{j_0, l}:l \in \mathbb{Z} \right\rbrace $ and $\left\lbrace x_{t,j_0, l}: l \in \mathbb{Z}\right\rbrace$, respectively. Similarly, the kernel and the field detail coefficient vectors, $\boldsymbol\beta_{j}$ and $\check{\mathbf{x}}_{t,j}$ contain all the coefficients $\left\lbrace \beta_{j,l} :l \in \mathbb{Z}\right\rbrace$ and $\left\lbrace  \check x_{t,j, l}:l \in \mathbb{Z}\right\rbrace$ respectively.

The vectors of the kernel and the field scaling and wavelet functions, $\boldsymbol\lambda$ and $\boldsymbol\mu$ respectively, are defined by
\begin{align}
&\boldsymbol\lambda^\top(s)=\left[
	\begin{array}{ccccc}
		 \boldsymbol\phi_{j_0}^\top(s) & \boldsymbol\psi_{j_0}^\top(s) & \boldsymbol\psi_{j_0+1}^\top(s) & \cdots & \boldsymbol\psi_{j}^\top(s)
		\end{array}
		\right] \label{KernelBasisVector} \\
&\boldsymbol\mu^\top (r)= \left[
			\begin{array}{ccccc}
				\boldsymbol\phi_{j_0}^\top(r) & \boldsymbol\psi_{j_0}^\top(r) & \boldsymbol\psi_{j_0+1}^\top(r) & \cdots & \boldsymbol\psi_{j}^\top(r) 
			\end{array} 
				\right]. 
\label{FieldBasisVector}
\end{align}
where $s=r-r'$. The vectors in  (\ref{KernelBasisVector}) and (\ref{FieldBasisVector}) are constructed in the same manner as the vectors in  (\ref{KernelWeights}) and (\ref{FieldWeights}).

To derive the state-space model, we first define the matrices 
\begin{align}\label{eq:Lambdax}
 \mathbf{\Lambda}_{x} &\triangleq \int_{\Omega}\boldsymbol\mu\left(r\right)\boldsymbol\mu^\top\left(r\right) dr \\
\label{eq:Lambdatheta}
 \mathbf{\Lambda}_{\theta} &\triangleq \int_{\Omega}\boldsymbol\mu\left(r\right) \int_\Omega { 
	   \boldsymbol\theta^\top\boldsymbol\lambda\left(r-r'\right)
	    \boldsymbol\mu^\top\left(r\right) 
	\, dr'dr}
\end{align}
Substituting \eqref{eq:KernelFiniteExpansion} and \eqref{eq:FieldFiniteExpansion} in \eqref{eq:DiscreteTimeModel}, pre-multipling by $\boldsymbol\mu\left(\mathbf r\right)$ and integrating over the space gives
\begin{align}\label{eq:DecomposedModel2} 
	&\mathbf{\Lambda}_{x} \mathbf{x}_{t+1}= 
	\xi\mathbf{\Lambda}_{x} \mathbf{x}_t +T_s \varsigma \mathbf{\Lambda}_{\theta}\mathbf{x}_t +\int_{\Omega}\boldsymbol\mu\left(r\right)e_t\left(r\right)dr.
\end{align}
Cross-multiplying by $\mathbf{\Lambda}_{x}^{-1}$ gives the state transition equation
\begin{align}\label{eq:StateEquation}
 \mathbf x_{t+1} &=\mathbf A(\boldsymbol \theta) \mathbf x_t+ \mathbf w_t\\
\label{eq:A_theta}
 \mathbf A(\boldsymbol \theta) &= T_s\varsigma\mathbf{\Lambda}_{x}^{-1}\mathbf{\Lambda}_{\theta}+\xi\mathbf I,
\end{align}
where $\mathbf I$ is the identity matrix. The disturbance becomes 
\begin{equation}\label{eq:Disturbance}
\mathbf w_t= \mathbf{\Lambda}_{x}^{-1}\int_{\Omega}\boldsymbol\mu \left(r\right)e_t\left(r\right)dr,
\end{equation}
which is a vector valued, zero-mean normally distributed white noise process with covariance (see \cite{Scerri2009} for proof)
\begin{equation}
\boldsymbol\Sigma_w =\mathbf{\Lambda}_{x}^{-1}\iint\limits_{\boldsymbol\Omega}\boldsymbol\mu\left(r\right) \eta\left(r-r'\right)\boldsymbol\mu^{\top}\left(r'\right)dr'dr\mathbf{\Lambda}_{x}^{-\top}.
\end{equation}
The observation equation of the state-space model is found by substituting decomposition \eqref{eq:FieldFiniteExpansion}
 into \eqref{eq:ObservationEquation} giving
\begin{equation}\label{eq:ReducedObservationEquation} 
	\mathbf{y}_t = \mathbf{C}\mathbf{x}_t + \boldsymbol{\varepsilon}_t,
\end{equation}
where each element of the observation matrix is given by
\begin{equation}
	\mathbf{C}_{ij} \triangleq \int_{\Omega}m(r_i - r')\boldsymbol{\mu}_j(r') \, d\mathbf{r}'.
\end{equation}

@@

where $m>1$, $\ast$ denotes convolution and $N_1\left(r\right)$ is the characteristic function of the unit interval $\left[ 0,1\right)$ ($rect$ function).

@@

where $\left\langle \cdot,\cdot\right\rangle $ denotes the inner product. This holds as $N_m\left(r\right)$ is isotropic. The $m^{th}$-order B-spline wavelet is defined as \cite{Chui1992}

@@

The state-space representation of the MRAIDE allows the use of the well known Expectation Maximization (EM) algorithm \cite{Dempster1977} to infer both the kernel and the field from electrophysiological data. The EM algorithm, when used in this context \cite{Dewar2009}, yields the maximum likelihood kernel estimate and the posterior distribution of the field over time. We use the standard EM algorithm for linear dynamic systems \cite{Roweis1999,Shumway2000} and hence only describe aspects of the algorithm that are specific to MRA neural field equations. EM essentially finds increasingly tighter lower bounds on the likelihood of the kernel $p(\mathbf{y}_1 \ldots \mathbf{y}_T|w)$ so that, at convergence, the maximum of the bound corresponds to the maximum of the likelihood. For the linear state-space model given by \eqref{eq:StateEquation} and \eqref{eq:ReducedObservationEquation}, the bound used is a quadratic of the form

@@

\parham{The actual connectivity kernel} and the decomposition is plotted in \figurename{\ref{fig:KernelEstimate}}(a). The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel lies inside the confidence interval. The small error in the estimate is attributed to the MRA of the system used to form the estimator. \figurename{\ref{fig:KernelEstimate}}(b) and (c) shows the kernel scaling and wavelet functions, respectively.
\begin{figure}[!h] 
	\centering
		\includegraphics[scale=1]{./Graph/Field.pdf}
		\caption{Actual (grey) and estimated (black) neural fields at a time instant for different spatial resolutions. (a) $j=0$, $n_x=17$, $RMSE = 2.10$. (b) $j=1$, $n_x=33$, $RMSE = 1.42$. (c) $j=2$, $n_x=65$, $RMSE = 0.80$. (d) $j=3$, $n_x=131$, $RMSE = 0.72$. (e) $j=4$, $n_x=263$, $RMSE=0.72$.}
	\label{fig:FieldEstimates}
\end{figure} 
\begin{figure}[!h] 
	\centering
% 		\includegraphics[scale=1]{./Graph/KernelUnknownSupportWithVarsigma.pdf}
		\includegraphics[scale=1]{./Graph/test.pdf}
		\caption{Connectivity kernel estimate. (a) The actual kernel is shown with the black solid line. The estimated kernel and confidence interval is shown by the red line and red shaded region ($\pm2$ std.). (b) Kernel scaling functions at $j=1$. (c) Kernel wavelet functions at $j=1$.}
	\label{fig:KernelEstimate}
\end{figure}
% \caption{The actual connectivity kernel is shown with the black solid line. The estimated kernel and confidence interval is shown by the red line and red shaded region ($\pm2$ std.). The 7 weighted kernel basis functions are shown by the black dashed lines.}
\section{Discussion}
In this paper, we have presented a novel model-based framework for estimating cortical dynamics from electrophysiological measurements. The novel and key developments of the paper include the multi-resolution \mike{FEM} representation of the neural field, and the estimator for an intracortical connectivity kernel with an arbitrary shape. This work is significant, as the ability to create patient specific neural field models has the potential to contribute to our understanding and improve treatment of diseases resulting from abnormal neurodynamics, such as epilepsy. Other groups have previously highlighted the importance of a multi-resolution approach in neural field modeling. For example, it is thought that the dynamics and connectivity structure differs at different spatial scales~\cite{Qubbaj2009}.

>>>>

Equations \eqref{eq:KernelExpansion} and \eqref{eq:FieldExpansion} are infinite series expansions and must be truncated at some level $j$. The field and the connectivity kernel are approximated by
\begin{align}
	w\left(r-r'\right) &\approx \boldsymbol\theta^\top\boldsymbol\lambda\left(r-r'\right) 
	\label{eq:KernelFiniteExpansion} \\
	v_t\left(r\right) &\approx \boldsymbol\mu^\top\left(r\right)\mathbf{x}_t,
	\label{eq:FieldFiniteExpansion}
\end{align}
where the unknown parameter and state vectors, $\boldsymbol\theta \in \mathbb{R}^{n_{\theta}}$ and $\mathbf{x}_t \in \mathbb{R}^{n_x}$, are defined as 
\begin{align}
\boldsymbol\theta^\top &=[\begin{array}{ccccc} \boldsymbol\alpha_{j_0}^\top & \boldsymbol\beta_{j_0}^\top & \boldsymbol\beta_{j_0+1}^\top & \cdots & \boldsymbol\beta_{j}^\top \end{array}] 
\label{KernelWeights} \\
\mathbf{x}_{t}^\top &=[\begin{array}{ccccc}\mathbf{x}_{t,j_{0}}^\top &  \check{\mathbf{x}}_{t,j_{0}}^\top & \check{\mathbf{x}}_{t,j_{0}+1}^\top & \cdots & \check{\mathbf{x}}_{t,j}^\top\end{array}].
\label{FieldWeights}
\end{align}
The kernel and field approximation coefficient vectors, $\boldsymbol \alpha_{j_0}$ and $\mathbf{x}_{t,j_{0}}$, contain all the coefficients $\left\lbrace\alpha_{j_0, l}:l \in \mathbb{Z} \right\rbrace $ and $\left\lbrace x_{t,j_0, l}: l \in \mathbb{Z}\right\rbrace$, respectively. Similarly, the kernel and the field detail coefficient vectors, $\boldsymbol\beta_{j}$ and $\check{\mathbf{x}}_{t,j}$ contain all the coefficients $\left\lbrace \beta_{j,l} :l \in \mathbb{Z}\right\rbrace$ and $\left\lbrace  \check x_{t,j, l}:l \in \mathbb{Z}\right\rbrace$ respectively.

The vectors of the kernel and the field scaling and wavelet functions, $\boldsymbol\lambda$ and $\boldsymbol\mu$ respectively, are defined by
\begin{align}
&\boldsymbol\lambda^\top(s)=[
	\begin{array}{ccccc}
		 \boldsymbol\phi_{j_0}^\top(s) & \boldsymbol\psi_{j_0}^\top(s) & \boldsymbol\psi_{j_0+1}^\top(s) & \cdots & \boldsymbol\psi_{j}^\top(s)
		\end{array}
		] \label{KernelBasisVector} \\
&\boldsymbol\mu^\top (r)= [
			\begin{array}{ccccc}
				\boldsymbol\phi_{j_0}^\top(r) & \boldsymbol\psi_{j_0}^\top(r) & \boldsymbol\psi_{j_0+1}^\top(r) & \cdots & \boldsymbol\psi_{j}^\top(r) 
			\end{array} 
				].
\label{FieldBasisVector}
\end{align}
where $s=r-r'$. The vectors in  (\ref{KernelBasisVector}) and (\ref{FieldBasisVector}) are constructed in the same manner as the vectors in  (\ref{KernelWeights}) and (\ref{FieldWeights}).

To derive the state-space model, we first define the matrices 
\begin{align}\label{eq:Lambdax}
 \mathbf{\Lambda}_{x} &\triangleq \int_{\Omega}\boldsymbol\mu\left(r\right)\boldsymbol\mu^\top\left(r\right) dr \\
\label{eq:Lambdatheta}
 \mathbf{\Lambda}_{\theta} &\triangleq \int_{\Omega}\boldsymbol\mu\left(r\right) \int_\Omega { 
	   \boldsymbol\theta^\top\boldsymbol\lambda\left(r-r'\right)
	    \boldsymbol\mu^\top\left(r\right) 
	\, dr'dr}
\end{align}
Substituting \eqref{eq:KernelFiniteExpansion} and \eqref{eq:FieldFiniteExpansion} in \eqref{eq:DiscreteTimeModel}, pre-multipling by $\boldsymbol\mu\left(\mathbf r\right)$ and integrating over the space gives
\begin{align}\label{eq:DecomposedModel2} 
	&\mathbf{\Lambda}_{x} \mathbf{x}_{t+1}= 
	\xi\mathbf{\Lambda}_{x} \mathbf{x}_t +T_s \varsigma \mathbf{\Lambda}_{\theta}\mathbf{x}_t +\int_{\Omega}\boldsymbol\mu\left(r\right)e_t\left(r\right)dr.
\end{align}
Cross-multiplying by $\mathbf{\Lambda}_{x}^{-1}$ gives the state transition equation
\begin{align}\label{eq:StateEquation}
 \mathbf x_{t+1} &=\mathbf A(\boldsymbol \theta) \mathbf x_t+ \mathbf w_t\\
\label{eq:A_theta}
 \mathbf A(\boldsymbol \theta) &= T_s\varsigma\mathbf{\Lambda}_{x}^{-1}\mathbf{\Lambda}_{\theta}+\xi\mathbf I,
\end{align}
where $\mathbf I$ is the identity matrix. The disturbance becomes 
\begin{equation}\label{eq:Disturbance}
\mathbf w_t= \mathbf{\Lambda}_{x}^{-1}\int_{\Omega}\boldsymbol\mu \left(r\right)e_t\left(r\right)dr,
\end{equation}
which is a vector valued, zero-mean normally distributed white noise process with covariance \parham{(see \cite{Scerri2009} for proof)}
\begin{equation}
\boldsymbol\Sigma_w =\mathbf{\Lambda}_{x}^{-1}\iint\limits_{\boldsymbol\Omega}\boldsymbol\mu\left(r\right) \eta\left(r-r'\right)\boldsymbol\mu^{\top}\left(r'\right)dr'dr\mathbf{\Lambda}_{x}^{-\top}.
\end{equation}
The observation equation of the state-space model is found by substituting decomposition \eqref{eq:FieldFiniteExpansion}
 into \eqref{eq:ObservationEquation} giving
\begin{equation}\label{eq:ReducedObservationEquation} 
	\mathbf{y}_t = \mathbf{C}\mathbf{x}_t + \boldsymbol{\varepsilon}_t,
\end{equation}
where each element of the observation matrix is given by
\begin{equation}\label{eq:Observationmatrix}
	\mathbf{C}_{ij} \triangleq \int_{\Omega}m(r_i - r')\boldsymbol{\mu}_j(r') \, d\mathbf{r}'.
\end{equation}

@@

where $m>1$, $\ast$ denotes convolution and $N_1\left(r\right)$ is the characteristic function of the unit interval $\left[ 0,1\right)$ \parham{($rect$ function, is different)}.

@@

where $\left\langle \cdot,\cdot\right\rangle $ denotes the inner product. \parham{This holds as the support of $N_m\left(r\right)$ is $\left[ 0,m\right]$ and  is symmetric with respect to $r=\frac{m}{2}$.} The $m^{th}$-order B-spline wavelet is defined as \cite{Chui1992}

@@

The state-space representation of the MRAIDE allows the use of the well known Expectation Maximization (EM) algorithm \cite{Dempster1977} to infer both the kernel and the field from electrophysiological data. The EM algorithm, when used in this context \cite{Dewar2009}, yields the maximum likelihood kernel estimate and the posterior distribution of the field over time. We use the standard EM algorithm for linear dynamic systems \cite{Roweis1999,Shumway2000} and hence only describe aspects of the algorithm that are specific to MRA neural field equations. EM essentially finds increasingly tighter lower bounds on the likelihood of the kernel $p(\mathbf{y}_1 \ldots \mathbf{y}_T|w)$ so that, at convergence, the maximum of the bound corresponds to the maximum of the likelihood. For the linear state-space model given by (\ref{eq:StateEquation}-\ref{eq:Observationmatrix}), the bound used is a quadratic of the form

@@

\parham{The actual connectivity kernel} and the decomposition is plotted in \figurename{\ref{fig:KernelEstimate}}(a). The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel lies inside the confidence interval. The small error in the estimate is attributed to the MRA of the system used to form the estimator. \figurename{\ref{fig:KernelEstimate}}(b) and (c) shows the kernel scaling and wavelet functions, respectively.
\begin{figure}[!h] 
	\centering
		\includegraphics[scale=1]{./Graph/fig1.pdf}
		\caption{Actual (grey) and estimated (black) neural fields at a time instant for different spatial resolutions. (a) $j=0$, $n_x=17$, $RMSE = 2.10$. (b) $j=1$, $n_x=33$, $RMSE = 1.42$. (c) $j=2$, $n_x=65$, $RMSE = 0.80$. (d) $j=3$, $n_x=131$, $RMSE = 0.72$. (e) $j=4$, $n_x=263$, $RMSE=0.72$.}
	\label{fig:FieldEstimates}
\end{figure} 
\begin{figure}[!h] 
	\centering
		\includegraphics[scale=1]{./Graph/fig2.pdf}
		\caption{Connectivity kernel estimate. (a) The actual kernel is shown with the black solid line. The estimated kernel and \mike{confidence interval} is shown by the red line and red shaded region ($\pm2$ std.). (b) Kernel scaling functions at $j=1$. (c) Kernel wavelet functions at $j=1$.}
	\label{fig:KernelEstimate}
\end{figure}
% \caption{The actual connectivity kernel is shown with the black solid line. The estimated kernel and confidence interval is shown by the red line and red shaded region ($\pm2$ std.). The 7 weighted kernel basis functions are shown by the black dashed lines.}
\section{Discussion}
In this paper, we have presented a novel model-based framework for estimating cortical dynamics from electrophysiological measurements. The novel and key developments of the paper include the multi-resolution \mike{FEM} representation of the neural field, and the estimator for an intracortical connectivity kernel with an arbitrary shape. This work is significant, as the ability to create patient specific neural field models has the potential to contribute to our understanding and improve treatment of diseases resulting from abnormal neurodynamics, such as epilepsy. Other groups have previously highlighted the importance of a multi-resolution approach in neural field modeling. For example, it is thought that the dynamics and connectivity structure differs at different spatial scales~\cite{Qubbaj2009}.
